---
title: "Stats Computing Project 1"
author: "Collin McGuire"
date: "9/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Statistical Computing Project 1
### Collin McGuire

##### Initial Setup

First, we load the tidyverse package. Then we load in the Tidy Tuesday date from January 21, 2020, which we then save as spotify_songs 

```{r Load Data}
library(tidyverse)
tuesdaydata <- tidytuesdayR::tt_load('2020-01-21') # Load the Tidy Tuesday data from 2020-01-21 and save as tuesdaydata
spotify_songs <- tuesdaydata$spotify_songs
```

##### Part 1: Explore Data

For this section, we will make use of the dplyr function to answer some exploratory data analysis questions about the Spotify song data


###### 1. How many songs are in each genre

The number of songs in each genre are displayed in the tibble below:

```{r Count}
library(dplyr)
songs <- as_tibble(spotify_songs)
songs <- arrange(songs, track_album_release_date) # Arrange by date
songs %>%
  count(playlist_genre, sort = TRUE)
```

###### 2.What is average value of energy and acousticness in the latin genre in this dataset?

The average values for energy and acousticsness of the latin genre are displayed in the tibble below:

```{r Average Value}
songs %>%
  filter(playlist_genre == "latin") %>%
  summarize(energy = mean(energy, na.rm = TRUE), acousticness = mean(acousticness, na.rm = TRUE))
```

###### 3.Calculate the average duration of song (in minutes) across all subgenres. Which subgenre has the longest song on average?

The genre with the longest song on average is New Jack Swing.

```{r Duration}
songs %>%
  group_by(playlist_subgenre) %>%
  summarize(duration_ms = mean(duration_ms/60000, na.rm = TRUE)) %>%
  arrange(desc(duration_ms))
```

###### 4. Make two boxplots side-by-side of the danceability of songs stratifying by whether a song has a fast or slow tempo. Define fast tempo as any song that has a tempo above its median value. On average, which songs are more danceable?

On average, slow songs are more danceable than fast songs.

```{r Boxplots}
# Save the median value as med, then create a new column,
# tempo_cat which takes on a value based on the value of med
med <- summarize(songs, tempo = median(tempo, na.rm = TRUE))
tempo_set <- songs %>%
  mutate(tempo_cat = case_when(
    tempo >= med$tempo ~ "Fast",
    tempo < med$tempo ~ "Slow"
  ))
# Create a base ggplot, then add a boxplot geom
base <- tempo_set %>%
          ggplot(aes(tempo_cat, danceability, fill = tempo_cat))
base + geom_boxplot()
```

##### Part 2: Convert nontidy data into tidy data

In this section, we will take a non-tidy dataset and turn it into tidy data.

###### 1. Select only unique distinct rows from the dataset based on the track_name and track_artist columns 

```{r Part2:1}
# Use the distinct function from the dplyr package to select
# unique combinations of track_name and track_artist
part2 <- distinct(spotify_songs,
                  track_name, track_artist,
                  .keep_all = TRUE)
```

###### 2/3. Add a new column called year_released listing just the year that the song was released. Keep only songs that were released in or after 1980

```{r Part2:2&3}
# Use the ymd function from the lubridate package to parse the year
# Use the year function to strip the date down to year
# Add a new column with the mutate function, and filter to newer than 1980
library(lubridate)
year_data <- ymd(part2$track_album_release_date, quiet = FALSE, tz = NULL, truncated = 2)
only_year <- year(year_data)
post_1980 <- part2 %>%
  mutate(year_released = only_year) %>%
  filter(year_released >= 1980)
```

###### 4. Add a new column with the duration of the song in minutes

```{r Part2:4}
# Add a new column by dividing the duration in ms by 60,000
post_1980 <- post_1980 %>%
  mutate(duration_in_minutes = duration_ms / 60000)
```

###### 5. For each year_released, calculate the mean of at least 6 of the audio features or descriptors

```{r Part2:5}
# Create average by year with group_by and summarize
average_by_year <- post_1980 %>%
  group_by(year_released) %>%
  summarize(loudness = mean(loudness, na.rm = TRUE),
            instrumentalness = mean(instrumentalness, na.rm = TRUE),
            danceability = mean(danceability, na.rm = TRUE),
            energy = mean(energy, na.rm = TRUE),
            speechiness = mean(speechiness, na.rm = TRUE),
            tempo = mean(tempo, na.rm = TRUE),)
```

###### 6. Convert this wide dataset into a long dataset with a new feature and mean_score column

```{r Part2:6}
# Use the pivot_longer command to create a tidy dataset
tidy_average_by_year <- average_by_year %>%
  pivot_longer(-year_released, names_to = "feature", values_to = "mean_score")
```

##### Part 3: Data visualization

In this part of the project, we will visualize the tidy dataset we created in Part 2.

```{r Part 3 Data Viz, fig.height = 8, fig.width = 12}
# Set a figure width, and a baseplot with year released and mean score
# Use the facet_wrap feature with the scales option to create a grid
# Add a lowess smoother with geom_smooth, titles with labs
# Use theme to adjust axis test, plot and panel backgrounds
baseplot <- ggplot(tidy_average_by_year, aes(x = year_released, y = mean_score, color = feature))
baseplot + geom_point() + 
  facet_wrap(feature ~ ., scales = "free") +
  geom_smooth() +
  labs(title = "Song Characteristics by Year", subtitle = "Characteristics such as energy, tempo, and instrumentalness all reached a nadir around the late 1990's and early 2000's before climbing again to a peak in \nthe 2010's. Loudness, danceability, and speechiness, however, all started off relatively lower than the other three characteristics, but steadily increased \nto their maximum over time" , caption = "Collin McGuire", x = "Year", y = "Average Characteristic Value") +
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust=1), axis.line = element_line(color = "black"), plot.background = element_rect(fill = "grey87"), panel.background = element_rect(fill = "ghostwhite"))
```

##### Part 4: Make the worst plot you can!

I want to apologize in advance for this plot. No one should have to stare at this image for any length of time.

```{r Part 4 Ugly Graph, fig.height = 8, fig.width = 12}
# Add fill and color with aes
# Use geom_hex as a scatterplot replacement
# Most changes are done in theme, and are quite ugly
p4base <- ggplot(post_1980, aes(x = track_popularity, y = year_released, fill = playlist_genre, color = playlist_subgenre))
p4base + geom_hex(alpha = 0.05) +
  geom_smooth() +
  ggtitle("Popularity and Year") +
  theme(axis.text.x = element_text(size = 36, angle = 85), title = element_text(size = 3), 
        legend.text = element_text(family = "HersheySymbol", face = "bold.italic"),
        plot.background = element_rect(fill = "darkmagenta"))


```

* The first modification I made to the graph was to increase the axis labels to an unreasonably large size, and then angle them at 85 degrees. Angling and enlarging the labels can be useful when the labels overlap, as in Part 5.

* The second modification I made was to have the fill and color represent different variables, which is extremely difficult to decipher on this figure. This modification might be useful if there were fewer categories and datapoints.

* The third modification I made was adjusting the alpha value to 0.05, making the hexagons almost transparent. This feature would be helpful for layered graphs, as it allows for seeing the layers underneath.

* The fourth modification I made was adding a lowess smoother line, which can be helpful to show trends in scatterplot. In this case, the line detracts from the figure, although there wasn't much to detract from in the first place.

* The fifth modification I made was making the titles of the graph very small. Adjusting the size of the titles could be useful to make them more visible.

* The sixth modification I made was changing the font of my legend to Greek symbols. This could be useful if you were performing math equations and needed Greek symbols.

* The seventh modification I made was changing the background to magenta. Changing the background of a plot can be useful to help the figure stand out against the background.


##### Part 5: Make my plot a better plot!

```{R Part 5, fig.height = 8, fig.width = 12}
# USe labs for the title
# Most adjustments also made in theme
spotify_songs %>% 
  ggplot(aes(y=track_popularity, x=playlist_subgenre, fill = playlist_genre)) +
  labs(title = "Track Popularity by Playlist Genre and Subgenre", x = "Playlist Subgenre", y = "Track Popularity") +
  geom_violin(trim = TRUE, scale = "count") +
  facet_wrap( ~ playlist_genre, scales = "free_x") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(size = 10, color = "black", angle = 30, hjust=1),
        panel.grid.major = element_line(color = "black"),
        axis.line = element_line(color = "ghostwhite"),
        legend.position = "none",
        plot.background = element_rect(fill = "ghostwhite"), panel.background = element_rect(fill = "grey90"))


```

* The first modification I made was changing the labels for each figure to a smaller size, and angling them, which makes them more readable

* The second modification I made was changing the axis labels from variable names to proper names, aiding readability

* The third modification I made was adding the scales option to the facet wrap, which allows the x axis to scale properly.

* The fourth modification I made was adding grid lines to the plots. While this detracts from the minimalism, it greatly improves the ability of a reader to track the level of each value.

* The fifth modification I made was removing the legend, as the information conveyed by the legend was already conveyed more effectively by the panel headings.

* The sixth modification I made was changing the violin plot scale from area to count, so that proportional number of observations could be observed from the figure.

* The seventh modification I made was centering the title above the plot, which helps with readability.
